Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-04-24 manuscript/manuscript.tex
ms updates

We conducted extensive simulations to understand how temporal covariance,
$G(t)$, and convergence correlations behave under (1) different quantitative
genetic fitness models, (2) different trait architectures (e.g. varying levels
of $V_A$ for fitness and the number of sites affecting fitness) and (3)
background selection. Furthermore, we use two population simulations to
investigate how convergence correlations depend on (1) the population sizes of
each selection line sampled from the main population, and (2) the direction the
trait is selected on in each line (i.e. in the same direction, differing
directions, or only one lines elected).

Due to the high computational burden of forward simulations, we modeled a
single 50 megabase region in a population of $N = 1000$ diploid individuals
with a neutral variant mutation rate of $10^{-8}$ and a recombination rate of
$10^{-8}$ per basepair. This is roughly analogous to a XXX chromosome of
\emph{Drosophila melanogaster}; however with this small population size and
mutation rate, the population mutation rate $\theta$ for the entire region
leads to far fewer neutral sites to calculate covariances and other statistics
on than expected for a region this length in \emph{D.  melanogaster}. Since our
main goal is to understand how the dynamics of statistics used in the paper and
how they are affected by background selection, fitness model, and trait
architecture, we use population frequencies rather than sampling frequencies.

All forward simulations were conducted using SLiM \parencite{Haller2019-vu},
and the simulation routines are available in the Github repository
\url{https://github.com/vsbuffalo/cvtk/}.

@@

We first investigated the effects of the selected trait's genetic architecture
on temporal covariances and $G(t)$ by neutrally burning in a population for
$10N$ generations, and selecting on the trait with an exponential fitness
function. By using an exponential fitness function (a form of selection
multiplicative across sites), we ensure there is no build up of LD due to
selection (XXX) and epistasis; this serves as the simplest directional
selection model of a trait to understand the effects of genetic architecture on
the statistics we have used in the paper.

\vb{TODO: note about different burnin pops?}

@@

@@

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as an lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
variance in allele frequency change (this is the difference between $G$ and
$G'$, the latter of which includes these variance terms \cite{Buffalo2019-io}).
To verify that $G(t)$ estimated every ten generations is indeed a lower bound,
we used a simulation procedure similar to the exponential fitness function
simulations (described in Supplementary Material Section
\ref{supp:genetic-arch}), and calculated the temporal covariances and $G(t)$
both each generation, and every ten generations. Unlike the simulations
described in \ref{supp:genetic-arch}, we began selection at $10N$ generations,
and used trait $V_A = 0.01$ and $L = 1000$ sites affecting the trait.

To further understand the effects of calculating temporal covariances every ten
generations rather than every generation, we also compared their magnitudes and
decay rates using the simulations described above. We find that ten generation
block temporal covariances are orders of magnitude larger but decay at similar
rates (see Supplementary Figure \ref{suppfig:supp-blocks}B; note the two y-axis
scales are different). The larger magnitude is expected, as each ten generation
block temporal covariance is the sum of XXX.

@@

\subsubsection{Convergence Correlations}

@@

  \caption{$G(t)$ trajectories (left) and temporal covariances (right) from
    truncation selection simulations for different numbers of individuals
    selected (line color). The dashed lines indicate the $G(t)$ trajectories
    calculated \emph{including} fixed sites, while the solid lines exclude
    fixed sites.

    both
  including fixed sites (solid lines) and not including fixed sites (dashed
lines). The dashed lines are identical to those in Supplementary Materials
Figure \ref{suppfig:supp-bgs-g}. As expected, including fixed sites shrinks the
$G(t)$ towards zero, and in some cases, they can become negative.}

>>>>

We conducted extensive simulations to understand how temporal covariance,
$G(t)$, and convergence correlations behave under (1) different quantitative
genetic fitness models, (2) different trait architectures (e.g. varying levels
of $V_A$ for fitness and the number of sites affecting fitness), (3) background
selection, and (4) different sampling periods. Furthermore, we use two
population simulations to investigate how convergence correlations depend on
(1) the population sizes of each selection line sampled from the main
population, and (2) the direction the trait is selected on in each line (i.e.
in the same direction, differing directions, or only one lines elected).

Due to the high computational burden of forward simulations over this wide
breadth of parameters, we modeled a single 50 megabase region in a population
of $N = 1000$ diploid individuals with a neutral variant mutation rate of
$10^{-8}$ and a recombination rate of $10^{-8}$ per basepair. This is roughly
analogous to a XXX chromosome of \emph{Drosophila melanogaster}; however with
this small population size and mutation rate, the population mutation rate
$\theta$ for the entire region leads to far fewer neutral sites to calculate
covariances and other statistics on than expected for a region this length in
\emph{D.  melanogaster}. Since our main goal is to understand how the dynamics
of statistics used in the paper and how they are affected by background
selection, fitness model, and trait architecture, we use population frequencies
rather than sampling frequencies.

All forward simulations were conducted using SLiM \parencite{Haller2019-vu} and
run and processed using Snakemake \parencite{Koster2012-iv}; all simulation
routines are available in the Github repository
\url{https://github.com/vsbuffalo/cvtk/}.

@@

We first investigated the effects of the selected trait's genetic architecture
on temporal covariances and $G(t)$ by neutrally burning in a population for
$10N$ generations, and selecting on the trait with an exponential fitness
function. By using an exponential fitness function (a form of selection
multiplicative across sites), we ensure there is no build up of LD due to
selection and epistasis \parencite{Turelli1990-kd,Bulmer1980-zo}; this serves
as the simplest directional selection model of a trait to understand the
effects of genetic architecture on the statistics we have used in the paper.

@@

%\vb{TODO: note about different burnin pops?}

\subsubsection{Genetic Architecture and Convergence Correlations}

Using the same exponential fitness function simulations described above, we
also investigated how the convergence correlation is impacted by (1) genetic
architecture, (2) the design of the selection experiment, e.g. how many
individuals are selected for each line from the founding population, and (3)
the direction of selection across the two populations ``lines". After burning
in $N = 1000$ diploid populations for $10N$ generations, we simulated two
equally-sized lines of sizes $n = \{50, 500, 1,000\}$ diploids, and imposed
three selection schemes across different simulation runs. First, we imposed a
convergent selection scheme, where the populations undergo exponential
directional selection in the same direction. We expect that the convergent
correlation under this convergent scheme should be positive, as the two lines
should share some haplotypes carrying beneficial alleles, and these are
selected in the same direction across the two lines. Second, we imposed
divergent selection, where the two lines again undergo exponential directional
selection, except in different directions. Here, we expect the convergence
correlation to be negative, as haplotypes that increase the selected trait in
one population are beneficial in the upward selected line, but deleterious in
the downward selected line. Third, we have a control selection scheme, where
one line is selected and the other is not; this is akin to the control line in
the \textcite{Castro2019-uk} study (see Figure \ref{fig:figure-2}C). In this
case, we expect to see no convergence correlation, as only one line is being
selected. Finally, across these two-line simulation studies, we expect that
smaller sampled line sizes should show weaker convergent correlations, as the
probability that the same haplotypes are selected between the two lines
decreases with smaller size.

  \caption{The convergence correlations across the two population line
    exponential directional selection simulations; panel rows are for differing
    line population sizes, and panel columns are the modes of selection across
    the lines (convergent, divergent, and only a single selected line control).
    Line color indicates the target genetic architecture, in number of loci
    affecting the trait's value. 95\% confidence intervals are also shown. Note
    that selection begins at generation five, which is the reference
    generation; this is indicated by the split in the lines.}

  \label{suppfig:convergence-corrs}
\end{figure}

Overall, our simulations confirm our hypotheses; see Supplementary Material
Figure \ref{suppfig:convergence-corrs}. We also find that in simulations where
we target a monogenic genetic architecture (i.e. the target number of
trait-affecting loci is $L=1$), the convergence correlations are generally much
weaker than those under a polygenic architecture. However, this effect is
mediated by the line population size; the difference in convergence correlation
between $L = 1$ and $L = 1000$ are more dissimilar when the line population
sizes are larger (compare the first column, last two rows). Overall, we find
that the convergence correlation is affected by both genetic architecture and
the size of the selected population lines.

@@

@@

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as an lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
variance in allele frequency change (this is the difference between $G(t)$ and
$G'(t)$, the latter of which includes these variance terms \cite{Buffalo2019-io}).
To verify that $G(t)$ estimated every ten generations is indeed a lower bound,
we used a simulation procedure similar to the exponential fitness function
simulations (described in Supplementary Material Section
\ref{supp:genetic-arch}), and calculated the temporal covariances and $G(t)$
both each generation, and every ten generations. Unlike the simulations
described in \ref{supp:genetic-arch}, we began selection at $10N$ generations,
and used trait $V_A = 0.01$ and $L = 1000$ sites affecting the trait.

To further understand the effects of calculating temporal covariances every ten
generations rather than every generation, we also compared their magnitudes and
decay rates using the simulations described above. We find that ten generation
block temporal covariances are orders of magnitude larger but decay at similar
rates (see Supplementary Figure \ref{suppfig:supp-blocks}B; note the two y-axis
scales are different). The larger magnitude is expected, as each ten generation
block temporal covariance is the sum of 45 temporal covariances between
adjacent generations (e.g. $10 \choose 2$).

@@

  \caption{$G(t)$ trajectories (left) and temporal covariances (right) from
    truncation selection simulations for different numbers of individuals
    selected (line color). Dashed lines indicate $G(t)$ trajectories and
    temporal covariances calculated \emph{including} fixed sites, while the
    solid lines exclude fixed sites. All values are averaged over 30 replicate
    simulations; the lines in the right figure are loess smoothed, while points
    are averages. The solid lines of the temporal covariances have been
    excluded in the left figure for clarity, but are similar except they do not
    become negative.}
